The Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-09-19 paper.tex
revisao da implementacao

@@

Internal events bring support for a limited form of subroutines, as depicted in 
Figure~\ref{lst.sub}.
The subroutine \code{inc} is defined as a loop (lines 3-6) that continuously 
awaits its identifying event (line 4), incrementing the value passed as 
reference (line 5).
A trail in parallel (lines 8-11) invokes the subroutine in reaction to event 
\code{A} through an \code{emit} (line 10).
Given the stacked execution for internal events, the calling trail pauses, the 
subroutine awakes (line 4), runs its body (yielding \code{v=2}), loops, and 
awaits the next ``call'' (line 4, again).
Only after this sequence that the calling trail resumes and passes the 
assertion test.

\begin{figure}
\begin{lstlisting}[numbers=left,xleftmargin=2em]
event int* inc; // subroutine `inc'
par/or do
    loop do     // definitions are loops
        var int* p = await inc;
        *p = *p + 1;
    end
with
    var int v = 1;
    await A;
    emit inc => &v; // call `inc'
    _assert(v==2);  // after return
end
\end{lstlisting}
\caption{ Subroutine \code{inc} is defined in a loop (lines 3-6), in parallel 
with the caller (lines 8-11).
\label{lst.sub}
}
\end{figure}

@@

The size of \code{state\_t} depends on the number of events in the application;
for an application with less than 253 events (plus the 3 states), one byte is 
enough.
%
The size of \code{label\_t} depends primarily on the number of \code{await} 
statements in the application---each \code{await} splits the code in two and 
requires a unique entry point in the code for its continuation.
Additionally, split \& join points for parallel compositions, \code{emit} 
continuations, and finalization blocks also require labels.
%
The \code{seqno} will eventually overflow during execution (every 256 
reactions).
However, given that the scheduler traverses all trails in each reaction, it can 
adjust them to properly handle overflows (actually 2 bits to hold the 
\code{seqno} is already enough).
%
The stack size depends on the maximum depth of nested emissions and is bounded 
to the maximum number of trails, e.g., a trail emits an event that awakes 
another trail, which emits an event that awakes another trail, and so on---the 
last trail cannot awake any trail, because they will be all hanged in a 
\code{STACKED} state.
%
In general, in constrained sensor network applications, the size of 
\code{trail\_t} is typically only 3 bytes (1 byte for each field), imposing a 
negligible overhead even for trails that only await a single event and 
terminate.

\subsubsection{Code Generation}

\begin{figure}[t]
\begin{minipage}[t]{0.31\linewidth}
\begin{lstlisting}
input void A;
event void e;
// TRAIL 0 - lbl Main
par/and do
  // TRAIL 0 - lbl Main
  await e;
  // TRAIL 0 - lbl Awake_e
  // TRAIL 0 - lbl And_chk
with
  // TRAIL 1 - lbl And_sub_2
  await A;
  // TRAIL 1 - lbl Awake_A_1
  emit e;
  // TRAIL 1 - lbl Emit_cont
  // TRAIL 1 - lbl And_chk
end
// TRAIL 0 - lbl And_out
await A;
// TRAIL 0 - lbl Awake_A_2
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.31\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
enum {
  Main = 1,  // ln  3
  Awake_e,   // ln  7
  And_chk,   // ln  8,15
  And_sub_2, // ln 10
  Awake_A_1, // ln 12
  Emit_cont, // ln 14
  And_out,   // ln 17
  Awake_A_2  // ln 19
};
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.36\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
while (<...>) {
  trail_t* trail = <...>
  switch (trail->lbl) {
  case Main:
    // activate TRAIL 1
    TRLS[1].evt = STACKED;
    TRLS[1].lbl = And_sub_2;
    TRLS[1].stk = cur_stack;

@@

    <...>  // other labels
  }
}
\end{lstlisting}
\end{minipage}
\caption{
%{\small %\textmd{
Static allocation of trails, entry-point labels, and generated code.
%}%}
\label{lst.impl.trails}
}
\end{figure}

The example in Figure~\ref{lst.impl.trails} illustrates how trails and labels 
are statically allocated in a program.
%
The program has a maximum of 2 trails, because the \code{par/and} (line 4) can 
reuse \emph{TRAIL 0}, and the join point (line 16) can reuse both \emph{TRAIL 
0} and \emph{TRAIL 1}.
%
Each label is associated with a unique identifier in the \code{enum}.
%
The static vector to hold the two trails in the example is defined as

\begin{lstlisting}
trail_t TRLS[2];
\end{lstlisting}

In the final generated $C$ code, each label becomes a \emph{switch case} 
working as the entry point to execute its associated code.
%
Figure~\ref{lst.impl.code} shows the corresponding code for the program of 
Figure~\ref{lst.impl.trails}.
%
The program is initialized with all trails set to \code{INACTIVE}.
Then, the scheduler executes the \emph{Main} label in \code{TRAIL-0}.
%
When the \emph{Main} label reaches the \code{par/and}, it ``stacks'' 
\code{TRAIL-1} (line 5--8) and proceeds to the code in \code{TRAIL-0} (lines 
10--15), respecting the deterministic execution order.
%
The code sets the running \code{TRAIL-0} to await \code{EVT\_e} on label 
\code{Awake\_e}, and then halts with a \code{break}.
%
The next iteration of the scheduler takes \code{TRAIL-1} and executes its 
registered label \code{ParAnd\_sub\_2} (lines 17-22), which sets \code{TRAIL-1} 
to await \code{EVT\_A} and also halts.

Regarding cancellation, trails in parallel are always allocated in subsequent 
slots in the static vector \code{TRLS}.
Therefore, when a \code{par/or} terminates, the scheduler sequentially searches 
and executes \code{FIN} trails within the range of the \code{par/or}, and then 
clears all of them to \code{INACTIVE} at once.
Given that finalization blocks cannot contain \code{await} statements, the 
whole process is guaranteed to terminate in bounded time.
Escaping a \code{loop} that contains parallel compositions also trigger the 
same finalization process.

TODO internal emit?

@@

As an example, Figure~\ref{lst.impl.tinyos} shows our binding for \emph{TinyOS} 
which maps \emph{nesC} callbacks to input events in \CEU.
%
The file \code{ceu.h} (included in line 3) contains all definitions for the 
compiled \CEU program, which are further queried through \code{\#ifdef}'s.
The file \code{ceu.c} (included in line 4) contains the main loop of \CEU 
pointing to the labels defined in the program.
The callback \code{Boot.booted} (lines 6-11) is called by TinyOS on mote 
startup, so we initialize \CEU inside it (line 7).
If the \CEU program uses timers, we also start a periodic timer (lines 8-10) 
that triggers callback \code{Timer.fired} (lines 13-17) every 10 milliseconds 
and advances the wall-clock time of \CEU (line 15)%
\footnote{We also offer a mechanism to start the underlying timer on demand to
avoid the ``battery unfriendly'' 10ms polling.}.
The remaining lines map pre-defined TinyOS events that can be used in \CEU 
programs, such as the light sensor (lines 19-23) and the radio transceiver 
(lines 25-36).

>>>>

TODO: names (other embedded languages)

@@

Internal events bring support for a limited form of subroutines, as depicted in 
Figure~\ref{lst.sub}.
The subroutine \code{inc} is defined as a loop (lines 3--6) that continuously 
awaits its identifying event (line 4), incrementing the value passed as 
reference (line 5).
A trail in parallel (lines 8--11) invokes the subroutine in reaction to event 
\code{A} through an \code{emit} (line 10).
Given the stacked execution for internal events, the calling trail pauses, the 
subroutine awakes (line 4), runs its body (yielding \code{v=2}), loops, and 
awaits the next ``call'' (line 4, again).
Only after this sequence that the calling trail resumes and passes the 
assertion test.

\begin{figure}
\begin{lstlisting}[numbers=left,xleftmargin=2em]
event int* inc; // subroutine `inc'
par/or do
    loop do     // definitions are loops
        var int* p = await inc;
        *p = *p + 1;
    end
with
    var int v = 1;
    await A;
    emit inc => &v; // call `inc'
    _assert(v==2);  // after return
end
\end{lstlisting}
\caption{ Subroutine \code{inc} is defined in a loop (lines 3--6), in parallel 
with the caller (lines 8--11).
\label{lst.sub}
}
\end{figure}

@@

The size of \code{state\_t} depends on the number of events in the application;
for an application with less than 253 events (plus the 3 states), one byte is 
enough.
%
The size of \code{label\_t} depends primarily on the number of \code{await} 
statements in the application---each \code{await} splits the code in two and 
requires a unique entry point in the code for its continuation.
Additionally, split \& join points for parallel compositions, \code{emit} 
continuations, and finalization blocks also require labels.
%
The \code{seqno} will eventually overflow during execution (every 256 
reactions).
However, given that the scheduler traverses all trails in each reaction, it can 
adjust them to properly handle overflows (actually 2 bits to hold the 
\code{seqno} is already enough).
%
The stack size depends on the maximum depth of nested emissions and is bounded 
to the maximum number of trails, e.g., a trail emits an event that awakes 
another trail, which emits an event that awakes another trail, and so on---the 
last trail cannot awake any trail, because they will be all hanged in a 
\code{STACKED} state.

In the context of embedded systems, the size of \code{trail\_t} is typically 
only 3 bytes (1 byte for each field), imposing a negligible memory overhead 
even for trails that only await a single event and terminate.
%
For instance, the \emph{CTP} collection protocol ported to \CEU has at most 
eight simultaneous lines of execution with an overhead of 2\% in comparison to 
the original version in \emph{nesC} (a dialect of \emph{C} for event-driven 
programming)~\cite{ceu.sensys13}.

\subsection{Code Generation and Scheduling}

\begin{figure}[t]
\begin{minipage}[t]{0.31\linewidth}
\begin{lstlisting}
input void A;
event void e;
// TRAIL 0 - lbl Main
par/and do
  // TRAIL 0 - lbl Main
  await e;
  // TRAIL 0 - lbl Awake_e
  // TRAIL 0 - lbl And_chk
with
  // TRAIL 1 - lbl And_sub_2
  await A;
  // TRAIL 1 - lbl Awake_A_1
  emit e;
  // TRAIL 1 - lbl Emit_cont
  // TRAIL 1 - lbl And_chk
end
// TRAIL 0 - lbl And_out
await A;
// TRAIL 0 - lbl Awake_A_2

.
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.31\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
enum {
  Main = 1,  // ln 3
  Awake_e,   // ln 7
  And_chk,   // ln 8,15
  And_sub_2, // ln 10
  Awake_A_1, // ln 12
  Emit_cont, // ln 14
  And_out,   // ln 17
  Awake_A_2  // ln 19
};

trail_t TRLS[2] = {
  { STACKED,  Main, 0 };
  { INACTIVE, 0,    0 };
};

.
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.37\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
void dispatch (trail_t* t) {
  switch (t->lbl) {
    case Main:
      // activate TRAIL 1
      TRLS[1].evt = STACKED;
      TRLS[1].lbl = And_sub_2;
      TRLS[1].stk = cur_stack;

@@

    <...>  // other labels
  }
}
\end{lstlisting}
\end{minipage}
\caption{
%{\small %\textmd{
Static allocation of trails, entry-point labels, and dispatch loop.
%}%}
\label{lst.impl.trails}
}
\end{figure}

In the final generated code in \emph{C}, each trail label representing an entry 
point becomes a \emph{switch case} with the associated code to execute.
%
Figure~\ref{lst.impl.trails} illustrates the generation process.
For the program in the left of the figure, the compiler extracts the entry 
points and associated trails.
%
The entry points translate to an \code{enum} in the generated code (lines 
1--10, in the middle of the figure).
The trails translate to a vector of \code{trail\_t} elements with the maximum 
number of simultaneous trails (lines 12--16).
Initially, \code{TRAIL-0} is set to execute the \code{Main} entry point, while 
all others are set to \code{INACTIVE}.

The scheduler executes in two passes:
%
In the \emph{broadcast} pass, the scheduler sets to \code{STACKED} all trails 
that are waiting for the current event to execute in the current stack level.
%
In the \emph{dispatch} pass, the scheduler executes each trail that is 
\code{STACKED} to run in the current level, setting it immediately to 
\code{INACTIVE}.

During the dispatch pass, if a trail executes and emits and internal event, the 
scheduler increments the stack level and re-executes the two passes.
After all trails are properly dispatched, the scheduler decrements the stack 
level and resumes the previous execution.
For the first reaction chain, the scheduler starts from the \emph{dispatch} 
pass, given that the \code{Main} label is the only one that can be active at 
the stack level 0 (line 13, in the middle of Figure~\ref{lst.impl.trails}).

The code the right of the Figure~\ref{lst.impl.trails} dispatches a trail
according to its current label to execute.
%
For the first reaction, it executes the \code{Main} label in \code{TRAIL-0}.
%
When the \code{Main} label reaches the \code{par/and}, it ``stacks'' 
\code{TRAIL-1} (lines 4--8) and proceeds to the code in \code{TRAIL-0} (lines 
9--14), respecting the deterministic execution order.
%
The code sets the running \code{TRAIL-0} to await \code{EVT\_e} on label 
\code{Awake\_e}, and then halts with a \code{break}.
%
The next iteration of \code{dispatch} takes \code{TRAIL-1} and executes its 
registered label \code{And\_sub\_2} (lines 16--21), which sets \code{TRAIL-1} 
to await \code{EVT\_A} and also halts.

Regarding abortion and finalization, when a \code{par/or} terminates, the 
scheduler makes a \emph{broadcast} pass for the \code{FIN} event, but limited 
to the range of trails covered by the terminating \code{par/or}.
Trails that do not match the \code{FIN} are set to \code{INACTIVE}, as they 
have to be aborted.
Given that trails in parallel are allocated in subsequent slots in the static 
vector \code{TRLS}, this pass only aborts the desirable trails.
The subsequent \emph{dispatch} pass executes the finalization code.
%Given that finalization blocks cannot contain \code{await} statements, the 
%whole process is guaranteed to terminate in bounded time.
Escaping a \code{loop} that contains parallel compositions also trigger the 
same abortion process.

@@

TODO: refs/cite WSNs

As an example, Figure~\ref{lst.impl.tinyos} shows our binding for \emph{TinyOS} 
which maps \emph{nesC} callbacks to input events in \CEU.
%
The file \code{ceu.h} (included in line 3) contains all definitions for the 
compiled \CEU program, which are further queried through \code{\#ifdef}'s.
The file \code{ceu.c} (included in line 4) contains the main loop of \CEU 
pointing to the labels defined in the program.
The callback \code{Boot.booted} (lines 6--11) is called by TinyOS on startup, 
so we initialize \CEU inside it (line 7).
If the \CEU program uses timers, we also start a periodic timer (lines 8--10) 
that triggers callback \code{Timer.fired} (lines 13--17) every 10 milliseconds 
and advances the wall-clock time of \CEU (line 15)%
\footnote{We also offer a mechanism to start the underlying timer on demand to
avoid the ``battery unfriendly'' 10ms polling.}.
The remaining lines map pre-defined TinyOS events that can be used in \CEU 
programs, such as the light sensor (lines 19--23) and the radio transceiver 
(lines 25--36).
